---
title: "Untitled"
output: html_document
date: "2023-11-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
library(tidyverse) 
library(sjPlot) 
library(tinytex) 
library(kableExtra)
library(car) 
library(knitr) 
library(ggplot2) 
library(emmeans) 
library(lme4) 
library(effects)
library(broom.mixed)
library(tidyr)
library(pbkrtest)
library(patchwork)
library(ICC)
library(glmmTMB)
library(HLMdiag)
library(lmeresampler)
library(psych)
library(bookdown)
library(tibble) #loading possibly relevant packages
```


```{r include=FALSE}
bmidata<- read.csv('~/Downloads/BMIbrainAge.csv')
demodata<- read.csv('~/Downloads/Demographics.csv')

#need to merge the demographic data into it as well, first delete repeated RIDs' 

#make sure 
demodata <- demodata[!duplicated(demodata$RID),] 
 
#then merge them 

merged<- merge(demodata,bmidata, by = "RID", all.y = TRUE)

#download the merged set onto csv then delete what i dont want in xl then re upload, easiest way to delete columns 
  
# Select RIDs with non-NA and non-9999 values in the cog beginning date column
selected_ridscog <- merged$RID[!is.na(merged$PTCOGBEG) & merged$PTCOGBEG != 9999]

# Print or use the selected RIDs as needed
print(selected_ridscog)

# Select RIDs with non-NA and non-9999 values in the mci beginning date column
selected_ridsmci <- merged$RID[!is.na(merged$PTMCIBEG) & merged$PTMCIBEG != 9999]

# Print or use the selected RIDs as needed
print(selected_ridsmci)

#  list of RIDs to remove
remove_rids <- c("4051", "4061", "4187", "4199", "4499", "4512", "4785","5207", "2315","2379")

# Remove participants with specified RIDs
mergedslim <- subset(merged, !(RID %in% remove_rids))


#download the data set to refine it in excel 
write.csv(mergedslim, file = "mergedslim.csv")

#changing from pounds to kilograms 1 = pounds

library(tidyverse)
merged_redu<- read.csv('~/Downloads/mergedslimref.csv')

merged_red1 <- merged_redu %>%
  mutate(
    WEIGHT = ifelse(WTUNIT == "1", round(WEIGHT * 0.45359237, 2), WEIGHT)
  )

#convert all heights to cm 
merged_red2<- merged_red1%>%
  mutate(
  HEIGHT = ifelse(HTUNIT == "1", round(HEIGHT*2.54,2), HEIGHT)
)

  #calculating BMI 

merged_red2<- merged_red2%>%
  mutate(
    HEIGHT_M2 = (HEIGHT/100)^2,
    BMI = round(WEIGHT/HEIGHT_M2,2)
  )

#remove 4376 no height data 
merged_red2<- subset(merged_red2, RID != 4376)

# Define the BMI ranges and corresponding labels
bmi_breaks <- c(-Inf, 18.5, 25, 30, Inf)
bmi_labels <- c("Underweight", "Healthy", "Overweight", "Obese")

# Create a new categorical variable based on BMI ranges
merged_red2$BMI_Group <- cut(merged_red2$BMI, breaks = bmi_breaks, labels = bmi_labels, include.lowest = TRUE)


#the data set i am now using is the merged_red2 dataset. It is complete now i will download it to compare to the data set before the height and weight were changed. 

#write the finalised data set before running the analysis and compare it to the old one to check it works
write.csv(merged_red2, file = "finaldataset2.csv")

finaldata<- merged_red2

#check BMI group is a factor 
is.factor(finaldata$BMI_Group)

# Use relevel to set "healthy" as the reference level
finaldata$BMI_Group <- relevel(finaldata$BMI_Group, ref = "Healthy")

finaldata<- subset(finaldata, BMI_Group != "Underweight")

regressiondata2 <- finaldata%>%
  select(- c("RESP","X", "PTCOGBEG","PTMCIBEG", "RESP","TEMP","TMPSRC"))

which(is.na(regressiondata2), arr.ind = TRUE)
#RID = 4485 remove from data set

regressiondata3<- subset(regressiondata2, RID != 4485)

#remove underweight we only have 3 participants 
levels(regressiondata3$BMI_Group)

# Remove level underweight 
regressiondata3$BMI_Group <- droplevels(regressiondata3$BMI_Group, exclude = "Underweight")

levels(regressiondata3$BMI_Group)

##check coding of factors and plausability of ranges
class(regressiondata3$SEX)
class(regressiondata3$PULSE)
class(regressiondata3$BPSYS)
class(regressiondata3$BPDIA)
class(regressiondata3$brainDelta)
class(regressiondata3$BMI_Group)


filter(regressiondata3, SEX > 2) 
filter(regressiondata3, SEX < 1) #no weird values


filter(regressiondata3, PULSE >100)
filter(regressiondata3, PULSE <60)

#there were a lot of ppt with low hr and one with faster HR, these conditions are common in the elderly and excluding them would not make sense, but it is something to note. these are the cut offs for brady and tachycardia 

range(regressiondata3$BPSYS)
range(regressiondata3$BPDIA)

#there is a wide range of blood pressures here with people showing extreme values with it being a significant predictor maybe run an extra analysis to see if there is an interaction between BMI and BP. 

#make a pulse pressure variable

regressiondata3<- regressiondata3%>%
  mutate(
    PULSEPRESS = BPSYS - BPDIA
  )

#make sex and racial category into a factor and rename the levels 
regressiondata3$SEX<- as.factor(regressiondata3$SEX)
regressiondata3$RACIAL_CAT<- as.factor(regressiondata3$RACIAL_CAT)

class(regressiondata3$SEX)
class(regressiondata3$RACIAL_CAT)

#rename the sex and racial categories
levels(regressiondata3$RACIAL_CAT) <- c("American Indian or Alaskan Native", "Asian", "Native Hawaiian or Other Pacific Islander", "Black or African American","White","More than one race","Unknown")
levels(regressiondata3$SEX)<- c("Male", "Female")

levels(regressiondata3$RACIAL_CAT)
levels(regressiondata3$SEX)

```




```{r exploring data, include=TRUE}
#Descriptive statistics 

desc_statistics <- regressiondata3 %>%
  group_by(BMI_Group) %>%
  summarise(
    "Average Brain Delta Score" = round(mean(brainDelta, na.rm = TRUE), 2),
    "SD of Brain Delta Score" = round(sd(brainDelta), 2), 
    "Lowest Scores on Brain Delta" = round(min(brainDelta),2),
   "Highest Scores on Brain Delta" = round(max(brainDelta),2),
   "Range of Brain Delta Scores" = round(max(brainDelta) - min(brainDelta),2),
   "pulse age avg" = round(mean(PULSEPRESS, na.rm = TRUE), 2),
   "pulse age Sd" = round(sd(PULSEPRESS), 2), 
   "Lowest Scores on pulse age" = round(min(PULSEPRESS),2),
   "Highest Scores on pulse age" = round(max(PULSEPRESS),2),
   "chron age avg" = round(mean(Age, na.rm = TRUE), 2),
   "chron age Sd" = round(sd(Age), 2), 
   "Lowest Scores on chron age" = round(min(Age),2),
   "Highest Scores on chron age" = round(max(Age),2),
  )


#descriptive graphs of bmi group and brain age delta score and blood pressure and brain age 

bdplot<- ggplot( data = regressiondata3, aes(x= BMI_Group, y = brainDelta, fill = BMI_Group))+
  geom_boxplot(alpha=0.3)+
  labs(x = "BMI Group", y = "BrainAGE Score")+
    scale_color_manual(values = c("cyan", "lightgreen", "lavender"))+
                         theme(
    text = element_text(family = "Times New Roman", size = 12))

plotBPD<- ggplot(data = regressiondata3, aes(x = BPDIA, y = brainDelta, col = BMI_Group))+
  geom_point()+
  geom_smooth(method = "lm", se = FALSE)+
  labs(x = "Diastolic Blood Pressure", y = "Brain Delta Score", col = "BMI Group")+
  scale_color_manual(values = c("red", "blue", "orange"))

plotBPs<- ggplot(data = regressiondata3, aes(x = BPSYS, y = brainDelta, col = BMI_Group))+
  geom_point()+
  geom_smooth(method = "lm", se = FALSE)+
  labs(x = "Systolic Blood Pressure", y = "Brain Delta Score", col = "BMI Group")+
    scale_color_manual(values = c("red", "blue", "orange"))


#BMI group bp interaction boxplots - not sure if needed but graphed

plotbmibpd<- ggplot(data = regressiondata3, aes(x = BMI_Group, y = BPDIA))+
  geom_boxplot()+
  geom_smooth(method = "lm", se = FALSE)+
  labs(x = "Diastolic Blood Pressure", y = "Brain Delta Score", col = "BMI Group")

plotbmibps<- ggplot(data = regressiondata3, aes(x = BMI_Group, y = BPSYS))+
  geom_boxplot()+
  geom_smooth(method = "lm", se = FALSE)+
  labs(x = "Systolic Blood Pressure", y = "Brain Delta Score", col = "BMI Group")

library(finalfit)
explanatory <- c("Age", "SEX", 
                "BMI", "BPSYS", 
                "BPDIA","BPSYS","PULSE","HEIGHT","WEIGHT","RACIAL_CAT", "PULSEPRESS")

dempot<- regressiondata3 %>% 
  summary_factorlist("BMI_Group", explanatory,
                     p=TRUE, na_include=FALSE,
                     add_dependent_label=TRUE,
                     dependent_label_prefix = "demo"
                     )

regressiondata3 <- regressiondata3 %>% 
  mutate(
    Age = ff_label(Age, "Age (years)"),
    SEX = ff_label(SEX, "Sex (Number of Participants)"),
    BPSYS = ff_label(BPSYS, "Systolic Blood Pressure (mmHg)"),
    BPDIA = ff_label(BPDIA, "Diastolic Blood Pressure (mmHg)"),
    PULSE = ff_label(PULSE, "Pulse Rate (bpm)"),
    HEIGHT = ff_label(HEIGHT, "Height (cm)"),
    WEIGHT = ff_label(WEIGHT, "Weight (kg)"),
    RACIAL_CAT = ff_label(RACIAL_CAT, "Racial Category (Number of Participants"),
    PULSEPRESS = ff_label(PULSEPRESS, "Pulse Pressure")
    )

table1 <- regressiondata3 %>%  
  summary_factorlist("BMI_Group", explanatory, 
                     p=TRUE, na_include=FALSE, 
                     add_dependent_label=TRUE,
                     dependent_label_prefix = "Demographics : "
                     )


bdplot
```
```{r}

pulsepressstats<- regressiondata3%>%
  group_by(BMI_Group)%>%
summarise(
"m" = mean(regressiondata3$PULSEPRESS),
"stans" = sd(regressiondata3$PULSEPRESS),
"range" = range(regressiondata3$PULSEPRESS),
"min" = min(regressiondata3$PULSEPRESS),
"max" = max(regressiondata3$PULSEPRESS)
)
```

```{r include=FALSE}
basedems<- regressiondata3%>%
  summarise(
    meanAge = mean(Age),
    Age_SD = sd(meanAge),
    Num_female = sum(SEX == "2"),
    minage = min(Age),
    maxage = max(Age)
    )

basedems

sdage<- regressiondata3%>%
  mutate(
    meaAGE = mean(Age),
    sdages = sd(Age)
  )

sdage
```

```{r}
write.table(model2, file = "modeloutput.txt", sep = ",", quote = FALSE, row.names = F)
```

```{r}
sum(regressiondata3$BMI_Group == "Obese")
sum(regressiondata3$BMI_Group == "Healthy")
sum(regressiondata3$BMI_Group == "Overweight")
```

```{r}
ggplot(regressiondata3, aes(x = BrainAge, y = BMI, color = BMI_Group)) +
  geom_point() +
  labs(x = "Brain Age", y = "BMI Scores", color = "BMI Group") +
  ggtitle("Scatter Plot of Brain Age vs BMI Scores") +
  theme_minimal()
```




#make a graph with the distribuion of chonological age and brain age overlayed on each other 

```{r model setup}
library(car)
newmodel<- lm(brainDelta~ SEX + PULSEPRESS + BMI_Group, data = regressiondata3)
summary(newmodel)
#plot it 
```



```{r}
plot(newmodel)
library(performance)
check_model(newmodel)
```

```{r include=FALSE}
#linearity
plot(newmodel, which = 1)
#independence
par(mfrow = c(1,2))
plot(resid(newmodel))
plot(fitted(newmodel), resid(newmodel))
#normality of errors 
plot(newmodel, which = 2)
#equal variances homoscedasticity 
residualPlots(newmodel)
plot(newmodel, which = 3)
#multicolinearity - values over 5 moderate correlation and over 10 strong correlation 
vif(newmodel)
#no values issue here
```

```{r}
mdl_diagnost <- 
  tibble(
  newmodel$model,
  fitted = fitted(newmodel),
  resid = residuals(newmodel),
  studres = rstudent(newmodel),
  hats = hatvalues(newmodel),
  cooksd = cooks.distance(newmodel)
)
```




```{r}
mdl_diagnost %>% 
  filter(abs(studres)>2)
#31 ppt

mdl_diagnost %>%
  filter(hats > (2*0.011))
#34 ppt

mdl_diagnost %>%
  filter(cooksd > 0.02) 

#33 ppt having a high influence - upon investigation there us only one person who has a cooks d that is way higher than the rest. It is also the person who has a pulse outside of the normal range. despite being in the healthy group. perform sensitivty analysis without this person. 
```


```{r}
car::influenceIndexPlot(newmodel, vars = "Cook",
                        id=F, main = "Cook's distance")
```

#sensitivity analysis 
```{r}
# 3 people over 0.02 remove them check the assumptions
reduceddata <- subset(regressiondata3, !(RID %in% c(4060, 5228, 6993)))
```

```{r}
#sensitivity analysis with ppt 120 removed due to high influence 
reducedmodel<- lm(brainDelta~ SEX + PULSEPRESS + BMI_Group, data = reduceddata)
summary(reducedmodel) 
```



```{r}
plot(reducedmodel)
car::influenceIndexPlot(reducedmodel, vars = "Cook",
                        id=F, main = "Cook's distance")
```
```{r}
mdl_diagnost2 <- 
  tibble(
  reducedmodel$model,
  fitted = fitted(reducedmodel),
  resid = residuals(reducedmodel),
  studres = rstudent(reducedmodel),
  hats = hatvalues(reducedmodel),
  cooksd = cooks.distance(reducedmodel)
)

mdl_diagnost2%>%
  filter(cooksd > 0.015) 
```

```{r}

# Gather summary statistics
stats.table <- as.data.frame(summary(reducedmodel)$coefficients)
# Get the confidence interval (CI) of the regression coefficient
CI <- confint(reducedmodel)
# Add a row to join the variables names and CI to the stats
stats.table <- cbind(row.names(stats.table), stats.table, CI)
# Rename the columns appropriately
names(stats.table) <- c("Term", "B", "SE", "t", "p", "CI_lower", "CI_upper")

stats.table
```

```{r}
library(flextable)
library(rempsyc)
my_table <- nice_table(stats.table)
```

```{r}

```


```{r}
intermodel<- lm(brainDelta~ BMI_Group*PULSEPRESS + SEX, data = reduceddata)
summary(intermodel)
#this is not good dont include this 
```



need to run power analysis and do demographics and tables for the results etc. 
the intercept is the normal group. they have a brain age delta of 

```{r}
#calculating mean average error and pearsons correlation 
mae<- corr.test(regressiondata3$Age, regressiondata3$BrainAge)
mae

```

```{r}
model2<- tab_model(reducedmodel)
```

```{r}
tab_model(reducedmodel, transform = NULL, auto.label = FALSE, collapse.ci=TRUE, file = "results.xls")
```

```{r}
library(psych)
corr.test(regressiondata3$PULSE, regressiondata3$BPDIA)
corr.test(regressiondata3$PULSE, regressiondata3$BPSYS)
corr.test(regressiondata3$BPDIA, regressiondata3$BPSYS)
```

```{r}
library(ggplot2)
ggplot(regressiondata3, aes(x = Age, fill = "Age")) +
  geom_density(alpha = 0.5) +
  geom_density(aes(x = BrainAge, fill = "Brain Age"), alpha = 0.5) +
  labs(x = "Chronological Age and Brain Age", y = "Distribution", ) +
  scale_fill_manual(name = "Type of Age",
                    values = c("Age" = "blue", "Brain Age" = "red"),
                    labels = c("Chronological Age", "Brain Age")) +
  theme( text = element_text(family = "Times New Roman", size = 12))
```

assumptions - run a levenes test 
need to plot the data to interpret
demographics of each group 



reconfigure inclusion of blood pressure - BMI blood pressure linked and the role of hypertension included it as a covariate - greatest bp greater brain age include in the discussion. BP might impact brain age for reasons that are different to obesity - further research the mech is not clear here suggest some. 

can include another model with obesity as 

need to know - plot dist of blood pressure scores in each of the groups 


we inckude the covariates becaue we havent age matched hten in the groups 


model bmi as cont with brain delta as well just for myself

tukey test to see between overweogh and obese

#You can compute the mean absolute error (MAE) either with the mae function in R (from the Metrics package) or by computing the average of the absolute value of the brain age delta column. You also report the model’s performance using the correlation between predicted age and chronological age, which you can also get from the BrainAgeResults file.

```{r}
library(Metrics)
maedata<- read.csv('~/Downloads/BrainAgeResults.csv')
actual<- maedata$Age
predicted<- maedata$BrainAge
mae(actual, predicted)
cor.test(maedata$Age, maedata$BrainAge)


```

```{r}
library(emmeans)
pairwise<- emmeans(reducedmodel, pairwise ~ BMI_Group)
summary(pairwise)
confint(pairwise)
```


```{r}
filter(regressiondata3,Age<60)
```

```{r}
newdataset<- subset(oldataset, ! = exclusion column names)
```





